### 2024-10-02 draft-2/draft-2.tex
material for states and transitions

This specification governs the interaction between two processes named the
{\it client} and the {\it engine}. \comment{Graphical interfaces, terminal
emulators, and scripts and utilities are examples of clients.} The text of
sections 1–5 is normative (except as described in {\it Conventions} below).

@@

\minor The engine's standard input and output must be open file descriptors and
the engine's standard error must be an open file descriptor until closed by the
engine. The sequence of bytes that the client sends to the engine via the
engine's standard input is the \term{client stream}. The sequence of bytes
that the engine sends to the client via the engine's standard output is the
\term{engine stream}.
\vskip-\parskip
\greybox{There are no requirements for an engine's standard error except that it
be open. For example, the engine's standard error may be directed to a null
file, to the client's standard output or standard error, or to a log file. There
are accordingly no restrictions on the sequence of bytes that the engine writes
to its standard error.}
\vskip-\parskip
\bluebox{No particular encoding is specified: the client and engine streams are
not required to be encoded sequences of Unicode scalar values (or code points
of any other character set). The specification instead governs the client and
engine streams as sequences of bytes {\it per se}.\par
However, clients and engines are recommended to use {\caps utf-8} for
client−engine communication so that engine, author, and option names are
displayed properly. Note that clients and engines may use different encodings
for other interfaces, such as for file system paths that are passed as arguments
to the operating system. For example, the client and engine may be communicating
over a network connexion and the client may be running on Linux (where paths
are arbitrary byte sequences that do not contain \hex{00} or \hex{2f}) but the
engine may be running on Windows (for which {\caps ntfs} is a common filesystem,
which stores file names in {\caps utf-16}, but where older libraries or
{\caps api}s may expect paths encoded in the local code page), and so the path
of a tablebase file may require transcoding, implicitly or explicitly.}

@@

\minor Message terminators divide the client and engine streams into
\term{messages}; that is, a message is a (possibly empty) subsequence of bytes
that does not contain a message terminator and every byte of the client and
engine streams is either part of a message or part of a message terminator.
The messages within the client stream are \term{client messages} and the
messages within the engine stream are \term{engine messages}.
\vskip-\parskip
\greybox{This precludes the proper use of some encodings, such as {\caps utf-16}
or {\caps utf-32} (since in {\caps utf-16} and {\caps utf-32}, the byte \hex{0a}
appears in the encodings of various scalar values, whereas in {\caps utf-8}, the
byte \hex{0a} only appears in the encoding for \uni{000a line feed}).}

@@

\vskip-0.5\parskip A nonnegative integer called the \term{depth from zeroing}.
\comment{This is the number of moves that have been played (none of which are
captures or pawn moves) since the last capture or pawn move, measured in ply.}

@@

@@

@@

\major{\bf Notation}

>>>>

This specification governs the interaction between two processes named the
{\it client} and the {\it engine}. \comment{Graphical interfaces, terminal
emulators, and scripts and utilities are examples of clients.} The text of
sections 1–4 is normative (except as described in {\it Conventions} below).

@@

\minor The engine's standard input and output must be open file descriptors and
the engine's standard error must be an open file descriptor until closed by the
engine. The sequence of bytes that the client sends to the engine via the
engine's standard input is the \term{client stream}. The sequence of bytes
that the engine sends to the client via the engine's standard output is the
\term{engine stream}.
\vskip-\parskip
\greybox{There are no requirements for an engine's standard error except that it
be open. For example, the engine's standard error may be directed to a null
file, to the client's standard output or standard error, or to a log file. There
are accordingly no restrictions on the sequence of bytes that the engine writes
to its standard error.}
\vskip-\parskip
\bluebox{No particular encoding is specified: the client and engine streams are
not required to be encoded sequences of Unicode scalar values (or code points
of any other character set). The specification instead governs the client and
engine streams as sequences of bytes {\it per se}.\par
However, clients and engines are recommended to use {\caps utf-8} for
client−engine communication so that engine, author, and option names are
displayed properly. Note that clients and engines may use different encodings
for other interfaces, such as for file system paths that are passed as arguments
to the operating system. For example, the client and engine may be communicating
over a network connexion and the client may be running on Linux (where paths
are arbitrary byte sequences that do not contain \hex{00} or \hex{2f}) but the
engine may be running on Windows (where {\caps ntfs} is a common filesystem,
which stores file names in {\caps utf-16}, but where older libraries or
{\caps api}s may expect paths encoded in the local code page), and so the
path of a tablebase file may require transcoding, implicitly or explicitly.}

@@

\minor Message terminators divide the client and engine streams into
\term{messages}; that is, a message is a (possibly empty) subsequence of bytes
that does not contain a message terminator and every byte of the client and
engine streams is either part of a message or part of a message terminator.
The messages within the client stream are \term{client messages} and the
messages within the engine stream are \term{engine messages}.
\vskip-\parskip
\greybox{This precludes the proper use of some encodings for client−engine
communication, such as \hbox{\caps utf-16} or \hbox{\caps utf-32} (since in
\hbox{\caps utf-16} and \hbox{\caps utf-32}, the byte \hex{0a} appears in the
encodings of various scalar values, whereas in \hbox{\caps utf-8}, the byte
\hex{0a} only appears in the encoding for \uni{000a line feed}).}
\vskip-\parskip
\greybox{Note that \hex{00} is not disallowed.}

@@

\vskip-0.5\parskip A nonnegative integer called the \term{depth from zeroing}.
\comment{This is the number of moves that have been played (none of which are
captures or pawn moves) since the last capture or pawn move, measured in~ply.}

@@

\minor {\caps todo}\hem a client message has one of seven types indicated
by the first token of the message (\s{debug}, \s{setoption}, \s{ucinewgame},
\s{position}, \s{isready}, \s{go}, and \s{stop})

@@

\minor {\caps todo}\hem an engine message has one of six types indicated
by the first token of the message (\s{id}, \s{option}, \s{uciok}, \s{info},
\s{readyok}, \s{bestmove})

@@

\minor Once the engine dequeues a \s{uci} message from the client stream,
at every point in time the engine is in one of six \term{states}. The engine
begins in the \term{initial} state. When the engine dequeues a client message
or enqueues an engine message, the state may change; this is a
\term{transition}.
\vskip-\parskip
\greybox{There is a particular invariant that transition rules should satisfy
due to the nature of standard input and standard output.

Firstly, the moment a message is read necessarily occurs some duration after
the moment that the message is written (that is, the sender does not directly
manipulate the memory of the recipient). Secondly, a process cannot both read
from a file descriptor and write to a file descriptor as an atomic transaction
(that is, “time of check to time of use” race conditions cannot be prevented).

In particular, it is impossible for a process to guarantee that a write
to file descriptor {\caps f} is not preceded by a write to file descriptor
{\caps g} by another process. Even if the process attempts to read {\caps g},
finds it empty, and immediately writes to {\caps f}, it is possible for an
intervening write to {\caps g} to occur before the write to {\caps f} occurs.

As a result, if a transition from state {\caps a} to state {\caps b} occurs
when the engine sends a message, it is impossible to guarantee that a message
sent by the client in state {\caps a} can actually be read by the engine in
state {\caps a} (and not in state {\caps b}). Therefore, any message that the
client is allowed to send in state {\caps a} should be allowed in
state~{\caps b}. The same principle holds vice versa, and so we can state the
invariant more generally as follows:

\it For distinct processes 1 and 2, for all pairs of states {\icap a} and
{\icap b}, if a transition from {\icap a} to {\icap b} can be caused by process
1 sending a message, every messages that process 2 is allowed to send in state
{\icap a} must also be allowed in state~{\icap b}.}

\minor In some states, the engine may dequeue a message. If the engine dequeues
a message, the message must be well-formed and have an allowed message type
given the state of the engine.
\vskip-0.5\baselineskip
\greybox{This is not meant to imply that standard input is ever required to be
empty, but meant to indicate that in some states the engine is not allowed to
read standard input.}

\minor In all states, the engine may enqueue a message. If the engine enqueues a
message, the message must be well-formed and have an allowed message type given
the state of the engine.

\minor The allowed message types that may be dequeued and enqueued in each state
are enumerated in the following table. If the {\caps next} column is not empty
for a particular row, the action listed in the row (dequeuing or enqueuing a
message of a particular type) causes a transition to the listed state.

\dimen0=4em
\dimen1=6em
\dimen2=6em
\dimen3=4em

\dimen4=0em
\advance\dimen4 by \dimen1
\advance\dimen4 by \dimen2

\def\br{\hrule height 0.8pt depth 0pt}
\def\hr{\hrule height 0.4pt depth 0pt}
\def\vr{\vrule width 0.4pt}

\vskip\baselineskip
\hbox to \hsize{%
  \hfil
  \vbox{%
    \br
    \hbox{\strut
      \hbox to \dimen0{\hfil\caps state\hfil}%
      \vr\hbox to \dimen4{\hfil\caps allowed\hfil}%
      \vr\hbox to \dimen3{\hfil\caps next\hfil}%
    \br
    \hbox{%
      \vtop{%
        \hbox{\strut}%
        \kern 0.4pt\hbox to \dimen0{\strut\hfil initial\hfil}%
      \vr
      \vtop{
        \hbox{\strut}%
        \kern 0.4pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
      \vtop{%
           \hbox to \dimen2{\strut\hem\s{id}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{option}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{uciok}\hfil}%
      \vr
      \vtop{%
        \hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \hr\hbox to \dimen3{\strut\hfil idle\hfil}%
    \br
    \hbox{%
      \vtop{%
        \hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \kern 0.4pt
        \kern 0.5\strutht\kern 0.5\strutdp
        \kern 0.2pt\hbox to \dimen0{\strut\hfil idle\hfil}%
      \vr
      \vtop{
        \hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \kern 0.4pt\hbox to \dimen1{\strut\hfil\caps dequeue\hfil}%
        \kern 0.4pt\hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \hr\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
      \vtop{%
           \hbox to \dimen2{\strut\hem\s{debug}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{setoption}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{ucinewgame}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{position}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{isready}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{go}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{stop}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{info}\hfil}%
      \vr
      \vtop{%
        \hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \hr\hbox to \dimen3{\strut\hfil sync\hfil}%
        \hr\hbox to \dimen3{\strut\hfil active\hfil}%
        \hr\hbox{\strut}%
    \br
    \hbox{%
      \vtop{%
        \hbox{\strut}%
        \kern -0.5\strutht\kern -0.5\strutdp
        \kern 0.2pt\hbox to \dimen0{\strut\hfil sync\hfil}%
      \vr
      \vtop{
        \hbox{\strut}%
        \kern -0.5\strutht\kern -0.5\strutdp
        \kern 0.2pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
      \vtop{%
        \hbox to \dimen2{\strut\hem\s{info}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{readyok}\hfil}%
      \vr
      \vtop{%
        \hbox{\strut}%
        \hr\hbox to \dimen3{\strut\hfil idle\hfil}%
    \br
    \hbox{%
      \vtop{%
        \hbox{\strut}%
        \kern 0.4pt\hbox{\strut}%
        \kern 0.4pt\hbox to \dimen0{\strut\hfil active\hfil}%
      \vr
      \vtop{
        \hbox{\strut}%
        \kern 0.4pt\hbox to \dimen1{\strut\hfil\caps dequeue\hfil}%
        \kern 0.4pt\hbox{\strut}%
        \hr
        \kern 0.5\strutht\kern 0.5\strutdp
        \kern 0.2pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
      \vtop{%
           \hbox to \dimen2{\strut\hem\s{debug}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{isready}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{stop}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{info}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{bestmove}\hfil}%
      \vr
      \vtop{%
        \hbox{\strut}%
        \hr\hbox to \dimen3{\strut\hfil ping\hfil}%
        \hr\hbox to \dimen3{\strut\hfil halt\hfil}%
        \hr\hbox{\strut}%
        \hr\hbox to \dimen3{\strut\hfil idle\hfil}%
    \br
    \hbox{%
      \vtop{%
        \hbox{\strut}%
        \kern -0.5\strutht\kern -0.5\strutdp
        \kern 0.2pt\hbox to \dimen0{\strut\hfil ping\hfil}%
      \vr
      \vtop{
        \hbox{\strut}%
        \kern -0.5\strutht\kern -0.5\strutdp
        \kern 0.2pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
      \vtop{%
        \hbox to \dimen2{\strut\hem\s{info}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{readyok}\hfil}%
      \vr
      \vtop{%
        \hbox{\strut}%
        \hr\hbox to \dimen3{\strut\hfil active\hfil}%
    \br
    \hbox{%
      \vtop{%
        \hbox{\strut}%
        \kern -0.5\strutht\kern -0.5\strutdp
        \kern 0.2pt\hbox to \dimen0{\strut\hfil halt\hfil}%
      \vr
      \vtop{
        \hbox{\strut}%
        \kern -0.5\strutht\kern -0.5\strutdp
        \kern 0.2pt\hbox to \dimen1{\strut\hfil\caps enqueue\hfil}%
      \vtop{%
        \hbox to \dimen2{\strut\hem\s{info}\hfil}%
        \hr\hbox to \dimen2{\strut\hem\s{bestmove}\hfil}%
      \vr
      \vtop{%
        \hbox{\strut}%
        \hr\hbox to \dimen3{\strut\hfil idle\hfil}%
    \br
  \hfil

{\caps todo}\hem pondering

\minor {\caps todo}\hem clients are required to send \s{ucinewgame} at least
once before the very first time they send \s{go} (but engines are strongly
recommended to behave identically whether or not \s{ucinewgame} is ever sent
or not)

\minor {\caps todo}\hem clients are required to send \s{position} at least
once before the very first time they send \s{go} (but engines are strongly
recommended to behave as if \s{position startpos} were sent if \s{position}
is not sent before \s{go} is sent)

{\caps todo}\hem note that the client is not required to send \s{isready}
before \s{go}

{\caps todo}\hem note that \s{ucinewgame} does not reset the current board
position nor does it reset the move history (only \s{position} changes the
current board state and move history)

{\caps todo}\hem note that the client may send \s{ucinewgame} after \s{position}
before \s{go} without any additional intervening \s{position}

{\caps todo}\hem note that engines and clients are not required to ignore
or otherwise handle unknown or unexpected tokens (these are simply violations)

% \major{\bf Notation}
